---
title: "Assignment 1"
author: "Alex Rintamaa"
date: '2024-03-01'
output: html_document
---

1) Consider a linear regression model. As we add additional parameters to our model (such as quadratic
terms or additional covariates), what is likely to happen to the expected values of the following? Briefly
explain your answers. (Note: our in-class example in week 1 looked at the observed bias and variance for
the training set. Here, you are answering about the expected bias and variance for the test set. You
should think back to the general rules about the bias-variance trade-off for this question.)
a) ğµğ‘–ğ‘ğ‘ [ğ‘“Ì‚ (ğ‘‹)]

flexibility increases while the expected bias decreases. This is a result because with less bias the number of predictors increases, which increases the number of Degrees of Freedom

b) ğ‘‰ğ‘ğ‘Ÿ[ğ‘“Ì‚ (ğ‘‹)]

flexibility will increase while the expected variance increases. This happens because the degrees of freedom increases as the number of predictors increase which is the case with increased expected variance

c) ğ‘‰ğ‘ğ‘Ÿ[ğœ–]

Because sigma^2 is a constant, the variance of the errors will not change based on flexibility. Because
errors are not effected by the number of predictors the, there cannot be a change bas on flexibility.

d) Training MSE

If the flexibility increases, the training MSE will decrease, The flexibility is increasing when training MSE decreases, as a result of more predictors being in the training MSE.

e) Test MSE

The test MSE will initially follow the same the training MSE, however as overfitting starts to happen the flexibility will start to decrease. This makes sense because it would follow the same trend as the expected bias flexibility.

2) Now consider a KNN model. As we increase the value of k, what is likely to happen to the expected
values of the following? Briefly explain your answers.

a) ğµğ‘–ğ‘ğ‘ [ğ‘“Ì‚ (ğ‘‹)]

Expected bias will increase with an increase in K, this can be proven by expected bias above, as an increase in K is equivalent to expected bias decreasing.

b) ğ‘‰ğ‘ğ‘Ÿ[ğ‘“Ì‚ (ğ‘‹)]

Expected variance will decrease with an increase in K, this can be proven by expected variance above, as an increase in K is equivalent to expected variance decreasing

c) ğ‘‰ğ‘ğ‘Ÿ[ğœ–]

As a result of Variance of the errors being constant, an increase in K would have no effect of the variance of errors.

d) Training MSE

Training MSE increases as a result of K increasing, this can be proven by the training MSE above, as an increase in K is equivalent to a decrease in training MSE.

e) Test MSE

Test MSE decreases as a result of K increasing, this can be proven by the testing MSE above, as an increase in K is equivalent to a increase in Test MSE.

3) Use the following R data frame as the training set for this problem.
# Training data for problem 3
knnData <- data.frame(X1 = c(5,3,6,9,4,7),
X2 = c(13,18,18,15,11,19),
Y = c(9,6,8,12,7,5))
a) Consider a new point (X1 = 6, X2 = 12). What are the Euclidean distances between this new point
and each of the 6 training data points?
b) What are the indices for the k=2 nearest neighbors to the new data point? What about the
indices for the k=3 nearest neighbors? Find the predicted Y value for both of these models.

```{r}
knnData <- data.frame(X1 = c(5,3,6,9,4,7),
 X2 = c(13,18,18,15,11,19),
 Y = c(9,6,8,12,7,5))

x <- data.frame(X1 = 6, X2 = 12)


euclid <- sqrt((knnData$X1 - x$X1)^2 + (knnData$X2 - x$X2)^2)
print("The Euclidean distances are")
print(euclid)

distance = order(euclid)
distance

knn2 = distance[1]
knn2
knn3 = distance[2]
knn3


# y_hat estimates
y_hat_k2 <- mean(knnData$Y[knn2])
y_hat_k3 <- mean(knnData$Y[knn3])


# Printing the predicted values.
print("The predicted values for y where k = 2 and y  where k = 3 are below.")
y_hat_k2
y_hat_k3
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(tidyverse)
library(MASS)        # lda and qda functions in MASS package
library(pdfCluster)  # package containing the olive oil data
library(ggpubr) # ggarrange()
```

## Read data in from website

```{r}
college <- read.csv("https://www.statlearning.com/s/College.csv",
row.names=1)
```

```{r}
# Standardize the numeric variables
college[,-1] <- as.data.frame(scale(college[,-1]))
```

```{r}
# Break into training and test data, DO NOT CHANGE THE SEED!
set.seed(02242023)
test_index <- sample(1:777, 277)
test_data <- college[test_index,]
train_data <- college[-test_index,]
```

4) Build a KNN model to classify schools as public or private using the number of full-time undergraduate
students (F.Undergrad) and the out-of-state tuition (Outstate), using the knn3() function in the
caret package. Try three different neighborhood sizes: k=1, k=10, and k=20. Compute the training and
test misclassification rates for each size. Which of these appears to be the best model for predicting
whether schools are public or private? Briefly explain.


## K = 1

```{r}
# 1-nearest neighbor
# Note the need to make mandate a factor! - matters with predict() later
k1_model <- knn3(as.factor(Private) ~ F.Undergrad + Outstate, data = train_data, k = 1)

# Predict test set using predict()
 # Note: Default is type="prob", make sure you specify class!
test_data$est.private.caret <- predict(k1_model, test_data, type ="class")
train_data$est.private.caret <- predict(k1_model, train_data, type = "class")


# Misclassification rate
misclass_rate <- mean(test_data$est.private.caret != test_data$Private)
misclass_train_rate <- mean(train_data$est.private.caret != train_data$Private)
misclass_rate
misclass_train_rate
```

## K = 10

```{r}
# 10-nearest neighbors
# Note the need to make mandate a factor! - matters with predict() later
k10_model <- knn3(as.factor(Private) ~ F.Undergrad + Outstate, data = train_data, k = 10)

# Predict test set using predict()
 # Note: Default is type="prob", make sure you specify class!
test_data$est.private.caret10 <- predict(k10_model, test_data, type ="class")
train_data$est.private.caret10 <- predict(k10_model, train_data, type = "class")

# Misclassification rate
misclass_rate2 <- mean(test_data$est.private.caret10 != test_data$Private)
misclass_train_rate2 <- mean(train_data$est.private.caret10 != train_data$Private)
misclass_rate2
misclass_train_rate2
```


## K = 20

```{r}
# 20-nearest neighbors
# Note the need to make mandate a factor! - matters with predict() later
k20_model <- knn3(as.factor(Private) ~ F.Undergrad + Outstate, data = train_data, k = 20)

# Predict test set using predict()
 # Note: Default is type="prob", make sure you specify class!
test_data$est.private.caret20 <- predict(k20_model, test_data, type ="class")
train_data$est.private.caret20 <- predict(k20_model, train_data, type = "class")


# Misclassification rate
misclass_rate3 <- mean(test_data$est.private.caret20 != test_data$Private)
misclass_train_rate3 <- mean(train_data$est.private.caret20 != train_data$Private)
misclass_rate3
misclass_train_rate3
```

The model that appears to be the best out of K=1, K=10, and K=20, would be the K=20, because this has the lowest misclass rate for both train and test. As the K increases it would appear that the misclass rate gets better.

5) Now, use a logistic regression model to classify schools as public or private using the same two
predictors (F.Undergrad and Outstate). Compute the training and test misclassification rates for this
classifier. How do these rates compare to the best KNN model in Problem 4?

### Logistic Regression

```{r}
log_college <- glm(as.factor(Private) ~ F.Undergrad + Outstate, family=binomial(link=logit), data=train_data)

# Predict test set using predict()
 # Note: Default is type="prob", make sure you specify class!
test_data$est.private.caretlog <- predict(log_college, test_data, type = "response")
train_data$est.private.caretlog <- predict(log_college, train_data, type = "response")

test_data$est.private.caretlog <- ifelse(test_data$est.private.caretlog > 0.5, "Yes", "No")
train_data$est.private.caretlog <- ifelse(train_data$est.private.caretlog > 0.5, "Yes", "No")


# Misclassification rate
misclass_ratelog <- mean(test_data$est.private.caretlog != test_data$Private)
misclass_train_ratelog <- mean(train_data$est.private.caretlog != train_data$Private)
misclass_ratelog
misclass_train_ratelog
```

The logistic regression model here is worse than the K=20 model for the the test misclassification rate. However, the training rate is not higher is slightly worse. Thus we defer to the test rate, saying that the k=20 model is better.

6) Lastly, use linear discriminant analysis to classify schools as public or private using the same two
predictors (F.Undergrad and Outstate).
a) As in Problems 4 and 5, compute the training and test misclassification rates for this classifier.
b) Compare these misclassification rates to the two other classifiers. Which classifier appears to be
the best classifier?
c) Create a scatterplot of number of full-time undergraduate students vs. out-of-state tuition, and
indicate the regions of the plot that are classified as public or private.


```{r}
# Creating lda model
lda_model <- lda(as.factor(Private) ~ F.Undergrad + Outstate, data = train_data)

# Predict test set using predict()
 # Note: Default is type="prob", make sure you specify class!
test_data$est.private.caretlda <- predict(lda_model, test_data)$class
train_data$est.private.caretlda <- predict(lda_model, train_data)$class


# Misclassification rate
misclass_ratelda <- mean(test_data$est.private.caretlda != test_data$Private)
misclass_train_ratelda <- mean(train_data$est.private.caretlda != train_data$Private)
misclass_ratelda
misclass_train_ratelda
```

The lda misclassification rates are much worse than both the K=20 model and the logistic regression model, having a higher rate than both of the other models by a good amount.

```{r}
ggplot(college, aes(x = F.Undergrad, y = Outstate, color = Private)) +
  geom_point() + 
  labs(x = "Full-Time Undergraduate Students",
       y = "Out-of-State Tuition",
       title = "Undergrads vs Out-of-State within Public Schools and Private Schools") +
  theme_bw()
```