---
title: "Homework 6"
author: "Alex Rintamaa"
date: "2024-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the library
library(pdfCluster)
library(caret)

# Open help menu
?oliveoil
# Load data for accessing
data("oliveoil")
```


1) Using 10-fold cross validation, fit and tune a support vector classifier for predicting the region of each
oil sample. Report any tuning parameters chosen, as well as the overall misclassification rate.


```{r}

# SVM Linear
olive.svc <- train(region ~ .,
             data =oliveoil,
             method = "svmLinear",
             trControl = trainControl("repeatedcv", number = 10, repeats = 10),
             tuneGrid = expand.grid(C = c(0.1, 1, 10, 100)))

olive.svc$bestTune
olive.svc$results

olive.misclass_svc<- 1 - max(olive.svc$results$Accuracy)
olive.misclass_svc

```
2) Using 10-fold cross validation, fit and tune a support vector machine with a polynomial kernel for
predicting the region of each oil sample. Report any tuning parameters chosen, as well as the overall
misclassification rate.


```{r}

# SVM Polynomial
olive.svmP <- train(region ~ .,
             data=oliveoil,
             method="svmPoly",
             trControl=trainControl("repeatedcv", number=10, repeats=10),
             tuneGrid=expand.grid(C=c(1,10,100),
                                 degree=2:4,
                                 scale=1))
olive.svmP$bestTune
olive.svmP$results

olive.misclass_svmP<- 1 - max(olive.svmP$results$Accuracy)
olive.misclass_svmP
```
3) Using 10-fold cross validation, fit and tune a support vector machine with a radial kernel for
predicting the region of each oil sample. Report any tuning parameters chosen, as well as the overall
misclassification rate.


```{r}

# SVM Radial
olive.svmR <- train(region ~ .,
              data = oliveoil,
              method = "svmRadial",
              trControl = trainControl("repeatedcv", number = 10, repeats = 10),
              tuneGrid = expand.grid(C = c(0.1, 1, 10, 100),
                                     sigma = c(0.01, 0.1, 1, 10)))
olive.svmR$bestTune
olive.misclass_svmR<- 1 - max(olive.svmR$results$Accuracy)
olive.misclass_svmR

```

4) Which model from 1-3 is best at classifying the regions? Display the confusion matrix for the best
model. Are there some regions where the classifier performs particularly well (with very few related
misclassifications)?


```{r}
# Compare SVM models using resampling
summary(resamples(list(linear = olive.svc,
                       poly = olive.svmP,
                       rad = olive.svmR)))

predictions_svmr <- predict(olive.svmR, newdata = oliveoil)

# Create the confusion matrix 
confusion_matrix_svmr <- confusionMatrix(predictions_svmr, oliveoil$region)
confusion_matrix_svmr
```


The best model we had was the SVM Radial Model, we can see that for some regions specifically Liguria East, Liguria West, Umbria, Apulia North, Sardinia Inland, and Sadinia Coast we are getting no false positves or false negatives.

2. 


```{r}
library(ggplot2)
# Define the curve equation
curve_eq <- function(X1, X2) {
  return((1 + X1)^2 + (2 - X2)^2 - 4)
}

# Generate data points for plotting
X1 <- seq(-3, 3, length.out = 100)
X2 <- seq(-1, 5, length.out = 100)
data <- expand.grid(X1 = X1, X2 = X2)
data$Class <- ifelse(curve_eq(data$X1, data$X2) > 0, "Outside", "Inside")



# Plot
ggplot(data, aes(x = X1, y = X2, color = Class, fill = Class)) +
  geom_point(size = 1.5) +
  geom_contour(aes(z = curve_eq(X1, X2)), breaks = 0, color = "black") +
  theme_minimal() +
  scale_color_manual(values = c("red", "blue"), name = "Class") +
  scale_fill_manual(values = c("red", "blue"), name = "Class") +
  labs(title = "Non-linear Decision Boundary") +
  theme(legend.position = "right")
```

c)
(0,0) has blue class classification
(-1,1) has red class classification
(2,2) has blue class classification
(3,8) has blue class classification

d)
By completing the square, you can get 

\[X_{1}^{2} + X_{2}^{2} + 2X_{1} - 4X_{2} + 1 = 0\]

To solve for any term in terms of a linear formula, it is quite easy.
