---
title: "Homework 2"
author: "Alex Rintamaa"
date: '2024-03-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(MASS) 
```

## Problem 1

1) (STA467: 5pt, STA567: 2.5pt) Explain the advantages of using cross-validation instead of validation
with a single test set?

The advantages of using cross-validation instead of validation with a single test set, is that more of the data is used in to trian the model as compared to validation with a single test set. Using more of the data to trian the model is important because it decreases the bias in the model. Cross-validation is also much better at predictiving overfitting.

## Problem 2

2) (STA467: 5pt, STA567: 2.5pt) What are the advantages and disadvantages of leave-one-out cross-
validation compared to k-fold cross-validation?

As a result of LOOCV estimates of test error being highly correlated with each other, the variance of the LOOCV estimate of test error is larger than the estimate of using k-fold cross-validation. However, for k-fold validation if we have a high bias, then the K-fold model is too simple, and is prone to underfitting. LOOCV is lias bias because it uses more data to train the model.

## Problem 3

3) (10pt) Use a 5-fold cross-validation to compare the following models for classifying the iris flowers by
species. NOTE: You should standardize your predictors, since we are using KNN.

```{r}
data(iris)
glimpse(iris)

standardize <- function(x){
  z <- (x - mean(x))/sd(x)
  return(z)
}

iris.standardized <- iris %>%
  mutate(SL_s = standardize(Sepal.Length),
         SW_s = standardize(Sepal.Width),
         PL_s = standardize(Petal.Length),
         PW_s = standardize(Petal.Width))
```
a) LDA with sepal length and sepal width
b) LDA with all 4 predictors
c) 5-nearest neighbors with all 4 predictors
d) 10-nearest neighbors with all 4 predictors
Report the cross-validation accuracy (i.e. 1 – MC rate) for each model:
Model Cross-Validation Accuracy

```{R}
set.seed(03082024)

add_cv_folds <- function(dat,cv_K){
  if(nrow(dat) %% cv_K == 0){ # if perfectly divisible, %%: mode, %/%: integer division
    dat$fold <- sample(rep(1:cv_K, each=(nrow(dat)%/%cv_K))) # shuffle
  } else { # if not perfectly divisible
    dat$fold <- sample(c(rep(1:(nrow(dat) %% cv_K), each=(nrow(dat)%/%cv_K + 1)), 
                              rep((nrow(dat) %% cv_K + 1):cv_K,
                                   each=(nrow(dat)%/%cv_K)) ) )
  }
  return(dat)
}

iris.standardized <- add_cv_folds(iris.standardized, 5)

kfold_preds1 <- rep(NA, 150)
kfold_preds2 <- rep(NA, 150)
kfold_preds3 <- rep(NA, 150)
kfold_preds4 <- rep(NA, 150)
kfold_acc1 <- rep(NA, 5)

for (j in 1:5) {
  # Separate our training (~400 obs) and test data (~100 obs)
  fold_index <- which(iris.standardized$fold == j)

  training <- iris.standardized[-fold_index, ]
  testing <- iris.standardized[fold_index, ]

  # 1) Train the model
  lda_spec <- lda(as.factor(Species) ~ SL_s + SW_s, data =training)
  lda_spec_4 <- lda(as.factor(Species) ~ SL_s + SW_s + PL_s + PW_s, data =training)
  knn5 <- knn3(as.factor(Species) ~ SL_s + SW_s + PL_s + PW_s, data=training, k=5)
  knn10 <- knn3(as.factor(Species) ~ SL_s + SW_s + PL_s + PW_s, data = training, k=10)

  # 2) Test predictions
  kfold_preds1[fold_index] <- predict(lda_spec, newdata = testing)$class
  kfold_preds2[fold_index] <- predict(lda_spec_4, newdata = testing)$class
  kfold_preds3[fold_index] <- predict(knn5, newdata = testing, type = "class")
  kfold_preds4[fold_index] <- predict(knn10, newdata = testing, type = "class")
  
}
```


# Accuracy Code and Answer

```{r}
# lda with 2 predictors accuracy and misclass
lda_mc <- mean(kfold_preds1 != as.numeric(iris.standardized$Species[1:150]))
acc_lda_mc <- 1 - lda_mc

# lda with all 4 predictors accuracy and misclass
lda_4_mc <- mean(kfold_preds2 != as.numeric(iris.standardized$Species[1:150]))
acc_lda_4_mc <- 1 - lda_4_mc

# knn4 accuracy and misclass
knn5_mc <- mean(kfold_preds3 != as.numeric(iris.standardized$Species[1:150]))
acc_knn5 <- 1 - knn5_mc

# knn10 accuracy and misclass
knn10_mc <- mean(kfold_preds4 != as.numeric(iris.standardized$Species[1:150]))
acc_knn10 <- 1 - knn10_mc

# creating accuracy data frame
acc_rate <- c(acc_lda_mc, acc_lda_4_mc, acc_knn5, acc_knn10)
acc_df <- data.frame(acc_rate)
acc_df
```


## Problem 4

4) (10pt) Suppose you are only interested in identifying I. versicolor. Create a binary variable that
identifies the species as versicolor or not. Use LOOCV to compare the following models for classifying iris
flowers as versicolor or not (using all 4 predictors):
a) Logistic regression
b) 10-nearest neighbors
c) LDA
For each model, compute the sensitivity, specificity, and the misclassification rate.

```{r, warning=FALSE}
iris.standardized1 <- iris %>%
  mutate(SL_s = standardize(Sepal.Length),
         SW_s = standardize(Sepal.Width),
         PL_s = standardize(Petal.Length),
         PW_s = standardize(Petal.Width),
         versI = ifelse(Species == "versicolor", "Yes", "No"))

# Initialize a vector for storing the predicted prices
loo_preds1 <- rep(NA, 150)
loo_preds2 <- rep(NA, 150)
loo_preds3 <- rep(NA, 150)

# LOOCV
for (j in 1:150) {
  # Separate our training (n-1 obs) and test data (1 obs)
  train_data <- iris.standardized1[-j,]
  test_data <- iris.standardized1[j,]
  
  # Fit model with training data (n-1 obs)
  log_vers <- glm(as.factor(versI) ~ SL_s + SW_s + PL_s + PW_s, family=binomial(link=logit), data=train_data)
  knn10_vers <- knn3(as.factor(versI) ~ SL_s + SW_s + PL_s + PW_s, data = train_data, k=10)
  lda_vers <- lda(as.factor(versI) ~ SL_s + SW_s + PL_s + PW_s, data = train_data)

  # Predict the remaining observation
  loo_preds1[j] <- predict(log_vers, test_data)
  loo_preds2[j] <- predict(knn10_vers, test_data, type = "class")
  loo_preds3[j] <- predict(lda_vers, test_data)$class
}
```

# Sensitivity, Specificity, and Misclassifcation Rate

```{r}
# changing the values for the log function.
loo_preds1 <- ifelse(loo_preds1 > 0.5, "Yes", "No")
# finding the log loocv misclassification rate
log_mc <- mean(loo_preds1 != iris.standardized1$versI[1:150])

# getting values for sensitivity and specificity calls
log_a <- sum(loo_preds1 == "Yes" & iris.standardized1$versI == "Yes")
log_b <- sum(loo_preds1 == "No" & iris.standardized1$versI == "Yes")  # false negative
log_c <- sum(loo_preds1  == "Yes" & iris.standardized1$versI == "No") # false positive
log_d <- sum(loo_preds1 == "No" & iris.standardized1$versI == "No")
# log sensitivity and specificity calls
log_sens <- log_a/(log_a + log_b)
log_spec <- log_d/(log_c + log_d)

# changing the value of iris.standard versI so comparison can be made
iris.standardized1$versI <- ifelse(iris.standardized1$versI == "Yes", 2, 1)

# finding the kNN loocv misclassification rate
knn10_loo <- mean(loo_preds2 != iris.standardized1$versI[1:150])

# getting values for sensitivity and specificity calls
knn10_a <- sum(loo_preds2 == 2 & iris.standardized1$versI == 2)
knn10_b <- sum(loo_preds2 == 1 & iris.standardized1$versI == 2)  # false negative
knn10_c <- sum(loo_preds2  == 2 & iris.standardized1$versI == 1) # false positive
knn10_d <- sum(loo_preds2 == 1 & iris.standardized1$versI == 1)

# kNN sensitivity and specificity calls
knn10_sens <- knn10_a/(knn10_a + knn10_b)
knn10_spec <- knn10_d/(knn10_c + knn10_d)

# finding the lda loocv misclassification rate
lda_loo_mc <- mean(loo_preds3 != iris.standardized1$versI[1:150])

# getting values for sensitivity and specificity calls
lda_a <- sum(loo_preds3 == 2 & iris.standardized1$versI == 2)
lda_b <- sum(loo_preds3 == 1 & iris.standardized1$versI == 2)  # false negative
lda_c <- sum(loo_preds3  == 2 & iris.standardized1$versI == 1) # false positive
lda_d <- sum(loo_preds3 == 1 & iris.standardized1$versI == 1)

# lda sensitivity and specificity calls
lda_sens <- lda_a/(lda_a + lda_b)
lda_spec <- lda_d/(lda_c + lda_d)

# creating columns for sensitivity, specificity, and misclassification rate
sens <- c(log_sens, knn10_sens, lda_sens)
spec <- c(log_spec, knn10_spec, lda_spec)
mc_rate <- c(log_mc, knn10_loo, lda_loo_mc)

answer <- data.frame(sens, spec, mc_rate)
answer
```


