---
title: "Homework 4"
author: "Alex Rintamaa"
date: "2024-04-09"
output: html_document
---

1. Explain the relationship between the number of knots and the degree of a spline regression model
and model flexibility.

 Too many knots can lead to overfitting, so it is important to find a balance. The regression spline is flexible in regions that contain a lot of knots as a result of those regions the polynomial coefficients change rapidly. When holding N as the number of knots, the degree of the spline regression model is N - 1. The degrees of freedom is N + D where D is the degree. For example, a cubic spline with 4 knots will have df = 5+3 = 8.

2. What are the spline basis functions for a cubic spline basis with 3 knots at values x1, x2, and x3?

$$
y_{i} = \beta_{0} + \beta_{1}x + \beta_{2}x^2 + \beta_{3}x^3 + h(x, x_{1}) + h(x, x_{2}) + h(x, x_{3}) + \epsilon_{i} \text{ where } h(x, x_{k}) = \{(x-x_{k}) \text{ if } x>x_{k} \text{ else } 0\} 

$$

```{r}
library(caret)
library(tidyverse)
library(MASS)
library(splines)

raceway <- read.table("raceway.txt", header = TRUE)
raceway
glimpse(raceway)
```

3) Fit the following models for predicting the top speed as a function of circumference. Report the leave-
one-out cross-validation error for each model. For all models that involve knots, use just one knot at 1.5.
(We would use more, but some of the models would become overparameterized with so few data
points). Which model best predicts the top speed? (12 points: 2pts each + 2pts comparison)
a) Linear Regression
b) Step Regression
c) Piecewise Linear Regression
d) Piecewise Polynomial Regression (degree=2)
e) Spline Regression (degree=2)

# Linear Regression

```{r, warning=FALSE}
lm_caret <- train(speed ~ circum,
                     method="lm",
                     data=raceway,
                     trControl=trainControl("loocv", savePredictions = TRUE),
                  preProcess = c("center", "scale"))
```
# Step Regression

```{r, warning=FALSE}
circum_knots <- c(min(raceway$circum), 1.5,
                   max(raceway$circum))

raceway$cut_circum <- cut(raceway$circum, breaks=circum_knots,
                               right=FALSE, include.lowest=TRUE)

step_caret <- train(speed ~ cut_circum,
                    method="lm",
                    data=raceway,
                    trControl=trainControl("loocv"),
                    preProcess=c("center", "scale"))
```

# Piecewise polynomial

```{r, warning=FALSE}
circum_knots <- c(min(raceway$circum), 1.5,
                   max(raceway$circum))

raceway$cut_circum <- cut(raceway$circum, breaks=circum_knots,
                               right=FALSE, include.lowest=TRUE)

piece_caret <- train(speed ~ cut_circum*circum,
                    method="lm",
                    data=raceway,
                    trControl=trainControl("loocv"),
                    preProcess=c("center", "scale"))
```

```{r, warning=FALSE}
circum_knots <- c(min(raceway$circum), 1.5,
                   max(raceway$circum))

raceway$cut_circum <- cut(raceway$circum, breaks=circum_knots,
                               right=FALSE, include.lowest=TRUE)

poly_caret <- train(speed ~ cut_circum*circum + cut_circum*I(circum^2),
                    method="lm",
                    data=raceway,
                    trControl=trainControl("loocv"),
                    preProcess=c("center", "scale"))
```

```{r, warning=FALSE}
circum_knots <- c(min(raceway$circum), 1.5,
                   max(raceway$circum))

raceway$cut_circum <- cut(raceway$circum, breaks=circum_knots,
                               right=FALSE, include.lowest=TRUE)


splines_caret <- train(speed ~ bs(circum, knots=1.5, degree=2),
                    method="lm",
                    data=raceway,
                    trControl=trainControl("loocv"),
                    preProcess = c("center", "scale"))

lm_caret$results
step_caret$results
piece_caret$results
poly_caret$results
splines_caret$results
```

```{r, warning = FALSE}
gam <- train(speed ~ bs(circum, knots = 1.5) + bs(banking, knots = 17),
             method = "lm",
             data=raceway,
             trContorol=trainControl("loocv"),
             preProcess = c("center", "scale"))
```


5.Fit a LOESS smoother to model the top speed as a function of circumference. Use a degree of 1, and
choose a value of span that makes the curve reasonably smooth. Plot the data and the smoother. (3
points)

```{r}

loess_caret <- train(speed ~ circum,
                     method="gamLoess",
                     data=raceway,
                     trControl=trainControl("cv", 5),
                     tuneGrid=data.frame(span=0.45, degree=1))
                     tuneGrid=expand.grid(span=seq(0.2,0.8,by=0.05), degree=1)  
loess_caret
smooth_data <- data.frame(circum = seq(min(raceway$circum), max(raceway$circum), length.out = 100))
smooth_data$preds <- predict(loess_caret, newdata = smooth_data)

ggplot() +
  geom_point(aes(x=circum, y=speed), data=raceway) +
  geom_line(aes(x=circum, y=preds), color="red", size=2, data=smooth_data) +
  geom_line(aes(x=circum, y=preds), color="blue", size=2, data=smooth_data) +
  theme_bw()
```
6. Fit a smoothing spline to model the top speed as a function of circumference. Try values of df from 1
to 5 by 0.25. Plot the data and the smoother. 

```{r}
smooth_caret <- train(speed ~ circum,
                  method="gamSpline",
                  data=raceway,
                  trControl=trainControl("cv", 5),
                  tuneGrid=data.frame(df=1:15))
smooth_caret

smooth_data <- data.frame(circum = seq(min(raceway$circum), max(raceway$circum), length.out = 100))
smooth_data$preds <- predict(smooth_caret,smooth_data)

ggplot() +
  geom_point(aes(x=circum, y=speed), data=raceway) +
  geom_line(aes(x=circum, y=preds), color="red", size=2, data=smooth_data) +
  geom_line(aes(x=circum, y=preds), color="blue", size=2, data=smooth_data) +
  theme_bw()
```